---
title: "RNA-seq workflow is the differential expression analysis using limma package"
author: "Dr Biruhalem"
date: "7/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##RNA-seq analysis in R

#https://combine-australia.github.io/RNAseq-R/06-rnaseq-day1.html

#https://master.bioconductor.org/packages/release/workflows/vignettes/RNAseq123/inst/doc/limmaWorkflow.html


```{r}
library(edgeR)
library(limma)
#library(org.Hs.eg.db)
library(gplots)
library(RColorBrewer)

library(dplyr)
```


#Reading in the data

```{r}

countdata <- BulkFrontalCortex3

head(countdata)

sampleinfo <- read.delim("sampleinfo2.txt", stringsAsFactors = TRUE)

head (countdata)
head(sampleinfo)


```

##Convert counts to DGEList object
Next we’ll create a DGEList object. This is an object used by edgeR to store count data. It has a number of slots for storing various parameters about the data.

```{r}
# create a DGEList object using the edgeR package:
y <- DGEList(countdata)

head(y$counts)

y$samples

names(y)

```


```{r}
table(colnames(countdata) == sampleinfo$SampName)
```




```{r}
#We can also store the groups for the samples in the DGEList object.

group <- paste(sampleinfo$Study)

group

# Convert to factor
group <- factor(group)

group



```

```{r}
# Add the group information into the DGEList
y$samples$group <- group
y$samples
```


# Filtering low expressed genes

```{r}

myCPM <- cpm(countdata)
head(myCPM)
thresh <- myCPM > 0.5
head(thresh)

# Summary of how many TRUEs there are in each row
# There are ... genes that have TRUEs in all samples.
table(rowSums(thresh))


# we would like to keep genes that have at least 2 TRUES in each row of thresh
keep <- rowSums(thresh) >= 2
summary(keep)


# Let's have a look and see whether our threshold of 0.5 does indeed correspond to a count of about 10-15
# We will look at the first sample
plot(myCPM[,1],countdata[,1])


# Let us limit the x and y-axis so we can actually look to see what is happening at the smaller counts
plot(myCPM[,1],countdata[,1],ylim=c(0,50),xlim=c(0,3))
# Add a vertical line at 0.5 CPM


y <- y[keep, keep.lib.sizes=FALSE]

head(y)

```


#Quality control, library size distribution


```{r}
# we can also adjust the labelling if we want
barplot(y$samples$lib.size/1e06, names=colnames(y), las=2, ann=FALSE, cex.names=0.75)
mtext(side = 1, text = "Samples", line = 4)
mtext(side = 2, text = "Library size (millions)", line = 3)
title("Barplot of library sizes")
```

#log2 transforming per million 

```{r}
# Get log2 counts per million
logcounts <- cpm(y,log=TRUE)
# Check distributions of samples using boxplots
boxplot(logcounts, xlab="", ylab="Log2 counts per million",las=2)
# Let's add a blue horizontal line that corresponds to the median logCPM
abline(h=median(logcounts),col="blue")
title("Boxplots of logCPMs (unnormalised)")

```

#varaible genes in the each groups 

```{r}
# Get the gene names for the top 500 most variable genes
select_var <- names(sort(var_genes, decreasing=TRUE))[1:500]
head(select_var)

# Subset logcounts matrix
highly_variable_lcpm <- logcounts[select_var,]
dim(highly_variable_lcpm)

head(highly_variable_lcpm)

```


#Normalisation for composition bias (TMM normalization)

```{r}
# Apply normalisation to DGEList object
y <- calcNormFactors(y, method = "TMM")

head(y)

y$samples
```

# Defferential analysis

```{r}
# Look at group variable again
group
```




```{r}
# Specify a design matrix without an intercept term
design <- model.matrix(~ 0 + group)
design
```

```{r}
## Make the column names of the design matrix a bit nicer
colnames(design) <- levels(group)
design
```


```{r}
#Voom transform the data

par(mfrow=c(1,1))

v <- voom(y,design,plot = TRUE)

v$targets

head(v$E)

```



```{r}
write.csv(v$E, file = "SSPETmmNormailizedvoom.csv")

```

```{r}
par(mfrow=c(1,2))

boxplot(logcounts, xlab="", ylab="Log2 counts per million",las=2,main="Unnormalised logCPM") 
abline(h=median(logcounts),col="blue")

## Let's add a blue horizontal line that corresponds to the median logCPM

boxplot(v$E, xlab="", ylab="Log2 counts per million",las=2,main="Voom transformed logCPM")
abline(h=median(v$E),col="blue")

```






```{r}
cont.matrix <- makeContrasts(SSPEvsControl=SSPE - GTEx, levels = design)
cont.matrix
```




```{r}

par(mfrow=c(1,2))

vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=cont.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")

```


```{r}
summary(decideTests(efit))
```

#including fold changes in the criteria 


```{r}
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)

```


```{r}

DE_SSPEvsControl_1 <- topTreat(efit, coef = 1, n=Inf)

head(DE_SSPEvsControl_1)


DE_SSPEvsControl <- topTreat(tfit, coef=1, n=Inf)

head(DE_SSPEvsControl)

#saving the whole differential expression analysis result

write.csv(DE_SSPEvsControl, "DE_SSPEvsControl.csv")

```


```{r}
#sorting by p-value 

ttable <- topTable(tfit,coef=1,sort.by="p")

head(ttable)
```


